---
title: "PSWQ scoping review: running the searches and initial findings"
author: "Magnus Johansson"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

This notebook runs the open-database searches for the PSWQ scoping review, caches
each raw result to `data/*.rds`, harmonizes the results to a common schema,
deduplicates, and summarizes what the searches return. It is a working document
and is deliberately kept out of the preregistration render (`_quarto.yml` renders
only `index.qmd`).

The programmatic searches here use Search A, the instrument-only query, which is
the primary strategy for recall. Method screening happens at the title and
abstract stage. As an initial finding, the OpenAlex results are also flagged for
whether a psychometric method is mentioned, which approximates the size of the
methods-restricted Search B. Lens.org, Dimensions, and CORE need a token or a
manual export and are read from `exports/` if files are present.

To re-run a search from scratch, delete its `.rds` in `data/` or pass
`refresh = TRUE` to `fetch_or_load()`.

## Setup

```{r}
#| label: setup

library(openalexR)
library(europepmc)
library(dplyr)
library(stringr)
library(tibble)
library(ggplot2)
library(httr2)

# Join OpenAlex's polite pool.
options(openalexR.mailto = "magnus.johansson.3@ki.se")

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_dir <- "data"
#cache_dir <- "data/dryrun_2026-09-23"
dir.create(cache_dir, showWarnings = FALSE)

# Load a cached .rds if present, otherwise evaluate `expr`, save it, return it.
# `expr` is only evaluated on a cache miss (lazy argument), so an existing cache
# never triggers a network call.
fetch_or_load <- function(name, expr, refresh = FALSE) {
  path <- file.path(cache_dir, paste0(name, ".rds"))
  if (!refresh && file.exists(path)) {
    message("loading cached ", path)
    return(readRDS(path))
  }
  result <- expr
  saveRDS(result, path)
  result
}

# Safe column access and DOI normalization.
col <- function(df, name) if (name %in% names(df)) df[[name]] else NULL
# First matching column among several candidate names. openalexR v2.0.0 renamed
# some columns (for example the abstract), so accept both old and new names.
col_any <- function(df, ...) {
  for (nm in c(...)) if (nm %in% names(df)) return(df[[nm]])
  NULL
}
norm_doi <- function(x) {
  x <- str_to_lower(as.character(x))
  x <- str_remove(x, "^https?://(dx\\.)?doi\\.org/")
  str_trim(x)
}

# Terms that indicate a psychometric method, used to flag Search-B-like records.
methods_pattern <- regex(
  paste(
    "rasch", "item response", "\\birt\\b", "graded response", "partial credit",
    "rating scale model", "factor analys", "confirmatory factor",
    "exploratory factor", "\\befa\\b", "\\bcfa\\b", "bifactor",
    "structural equation", "measurement invariance",
    "differential item functioning", "\\bdif\\b", "dimensional", "psychometric",
    sep = "|"
  ),
  ignore_case = TRUE
)

# The instrument-only query, expressed per source below.
instrument_terms <- '"Penn State Worry Questionnaire" OR PSWQ'

# Europe PMC has no combined title/abstract field, and a non-field-qualified
# query also matches full text for open-access articles, which inflates counts.
# Restrict to title and abstract explicitly. Phrases carry their own quotes.
epmc_tiab <- function(terms) {
  paste(sprintf('(TITLE:%s OR ABSTRACT:%s)', terms, terms), collapse = " OR ")
}
instrument_epmc <- epmc_tiab(c('"Penn State Worry Questionnaire"', 'PSWQ'))
methods_terms <- c(
  'Rasch', '"item response theory"', 'IRT', '"graded response model"',
  '"partial credit model"', '"rating scale model"', '"factor analysis"',
  '"confirmatory factor analysis"', '"exploratory factor analysis"',
  'bifactor', '"structural equation modeling"', '"structural equation modelling"',
  '"measurement invariance"', '"differential item functioning"', 'psychometric*'
)
methods_epmc <- epmc_tiab(methods_terms)

# Europe PMC's lite search (europepmc package) returns no abstracts. Query the
# REST API with resultType=core to get abstractText in bulk, paginating by
# cursorMark.
epmc_core_search <- function(query, page_size = 1000) {
  base <- "https://www.ebi.ac.uk/europepmc/webservices/rest/search"
  out <- list(); cursor <- "*"
  repeat {
    resp <- request(base) |>
      req_url_query(query = query, resultType = "core", format = "json",
                    pageSize = page_size, cursorMark = cursor) |>
      req_retry(max_tries = 3) |>
      req_perform() |>
      resp_body_json()
    hits <- resp$resultList$result %||% list()
    out  <- c(out, hits)
    nxt  <- resp$nextCursorMark
    if (length(hits) == 0 || is.null(nxt) || identical(nxt, cursor)) break
    cursor <- nxt
  }
  out
}
```

## OpenAlex

```{r}
#| label: openalex

oa_raw <- fetch_or_load("openalex_A", oa_fetch(
  entity = "works",
  from_publication_date = "2010-01-01",
  title_and_abstract.search = instrument_terms,
  verbose = FALSE
))

tidy_oa <- function(df) {
  date <- col_any(df, "publication_date")
  yr   <- col_any(df, "publication_year") %||%
    (if (!is.null(date)) as.integer(substr(as.character(date), 1, 4)) else NA_integer_)
  tibble(
    source   = "OpenAlex",
    doi      = norm_doi(col_any(df, "doi") %||% NA_character_),
    title    = as.character(col_any(df, "title", "display_name") %||% NA_character_),
    year     = suppressWarnings(as.integer(yr)),
    abstract = as.character(col_any(df, "abstract", "ab") %||% NA_character_)
  )
}

oa_tidy <- tidy_oa(oa_raw)
nrow(oa_tidy)
```

1021

## Europe PMC

Restricted to title and abstract via `instrument_epmc`, because a non-field
query also matches full text for open-access articles and inflates the count.
This keeps Europe PMC comparable to the OpenAlex and PubMed title/abstract
searches. Full-text-only mentions that are dropped are papers that cite a PSWQ
study rather than use the instrument, and citation chasing recovers any that
matter. The REST API is queried with `resultType=core`, so abstracts come back
too, which lets the methods-mention flag below cover Europe PMC as well as
OpenAlex.

```{r}
#| label: europepmc

epmc_raw <- fetch_or_load("europepmc_A_core", epmc_core_search(
  paste0('(', instrument_epmc, ') AND (PUB_YEAR:[2010 TO 2026])')
))

epmc_tidy <- tibble(
  source   = "Europe PMC",
  doi      = norm_doi(vapply(epmc_raw, function(x) x$doi %||% NA_character_, character(1))),
  title    = vapply(epmc_raw, function(x) x$title %||% NA_character_, character(1)),
  year     = vapply(epmc_raw, function(x) suppressWarnings(as.integer(x$pubYear %||% NA)), integer(1)),
  abstract = vapply(epmc_raw, function(x) x$abstractText %||% NA_character_, character(1))
)
nrow(epmc_tidy)
```

426.

## PubMed

Wrapped so a network or parsing hiccup does not stop the notebook. Leaves no
`pm_tidy` object if it fails, and the combine step below simply skips it.

```{r}
#| label: pubmed

pm_raw <- tryCatch(
  fetch_or_load("pubmed_A", {
    library(rentrez)
    term <- paste0('("Penn State Worry Questionnaire"[tiab] OR PSWQ[tiab]) ',
                   'AND 2010:2026[pdat]')
    # Use the web history so summaries are fetched by POST in batches. Passing
    # many ids to entrez_summary as a GET overflows the URL (HTTP 414).
    srch <- entrez_search(db = "pubmed", term = term, retmax = 0,
                          use_history = TRUE)
    recs <- list()
    if (srch$count > 0) {
      step <- 200
      for (start in seq(0, srch$count - 1, by = step)) {
        chunk <- entrez_summary(db = "pubmed", web_history = srch$web_history,
                                retstart = start, retmax = step,
                                always_return_list = TRUE)
        recs <- c(recs, chunk)
        Sys.sleep(0.34)   # stay under 3 requests/second without an API key
      }
    }
    recs
  }),
  error = function(e) { message("PubMed skipped: ", conditionMessage(e)); NULL }
)

if (!is.null(pm_raw) && length(pm_raw) > 0) {
  recs <- if (!is.null(pm_raw$uid)) list(pm_raw) else pm_raw   # single vs many
  get_doi <- function(x) {
    ids <- x$articleids
    d <- if (is.data.frame(ids)) ids$value[ids$idtype == "doi"] else character(0)
    if (length(d)) d[1] else NA_character_
  }
  pm_tidy <- tibble(
    source   = "PubMed",
    doi      = norm_doi(vapply(recs, get_doi, character(1))),
    title    = vapply(recs, function(x) x$title %||% NA_character_, character(1)),
    year     = vapply(recs, function(x) suppressWarnings(as.integer(substr(x$pubdate, 1, 4))), integer(1)),
    abstract = NA_character_
  )
  nrow(pm_tidy)
}
```

403.

## Semantic Scholar

```{r}
#| label: semanticscholar

s2_raw <- tryCatch(
  fetch_or_load("semanticscholar_A", {
    library(httr2)
    fetch_page <- function(token = NULL) {
      request("https://api.semanticscholar.org/graph/v1/paper/search/bulk") |>
        req_url_query(
          query  = '"Penn State Worry Questionnaire" | PSWQ',
          year   = "2010-",
          fields = "title,year,externalIds",
          token  = token
        ) |>
        req_retry(max_tries = 3) |>
        req_perform() |>
        resp_body_json()
    }
    out <- list(); tok <- NULL
    repeat {
      pg  <- fetch_page(tok)
      out <- c(out, pg$data %||% list())
      tok <- pg$token
      if (is.null(tok)) break
    }
    out
  }),
  error = function(e) { message("Semantic Scholar skipped: ", conditionMessage(e)); NULL }
)

if (!is.null(s2_raw) && length(s2_raw) > 0) {
  s2_tidy <- tibble(
    source   = "Semantic Scholar",
    doi      = norm_doi(vapply(s2_raw, function(x) x$externalIds$DOI %||% NA_character_, character(1))),
    title    = vapply(s2_raw, function(x) x$title %||% NA_character_, character(1)),
    year     = vapply(s2_raw, function(x) suppressWarnings(as.integer(x$year %||% NA)), integer(1)),
    abstract = NA_character_
  )
  nrow(s2_tidy)
}
```

825.

## Search B: methods-restricted counts

Search A above returns every record that mentions the PSWQ, most of which use the
instrument without evaluating it. The methods-restricted Search B is the count
that sets the screening load. Europe PMC runs it as an API query. OpenAlex is
approximated by filtering its Search A abstracts.

```{r}
#| label: searchB

epmcB <- fetch_or_load("europepmc_B_ta", epmc_search(
  query = paste0('(', instrument_epmc, ') AND (', methods_epmc,
                 ') AND (PUB_YEAR:[2010 TO 2026])'),
  limit = 3000, output = "parsed"
))

# OpenAlex Search A returns abstracts, so filter those locally for a real count.
oaB_n <- sum(str_detect(paste(oa_tidy$title, oa_tidy$abstract), methods_pattern),
             na.rm = TRUE)

tibble::tibble(
  source    = c("OpenAlex (local abstract filter)", "Europe PMC (title/abstract)"),
  n_methods = c(oaB_n, nrow(epmcB))
)
```

  source                           n_methods
  <chr>                                <int>
1 OpenAlex (local abstract filter)       161
2 Europe PMC (title/abstract)             70


## Manual exports (Lens, Dimensions, CORE)

Drop RIS or BibTeX exports from Lens.org, Dimensions, or CORE into `exports/`.
This reads any that are present with `synthesisr`. Skipped when the folder is
empty.

```{r}
#| label: exports

exp_files <- list.files("exports", pattern = "\\.(ris|bib|txt)$",
                        full.names = TRUE, ignore.case = TRUE)

if (length(exp_files) > 0 && requireNamespace("synthesisr", quietly = TRUE)) {
  exp_raw <- synthesisr::read_refs(exp_files)
  exp_tidy <- tibble(
    source   = "Manual export",
    doi      = norm_doi(col(exp_raw, "doi") %||% NA_character_),
    title    = as.character(col(exp_raw, "title") %||% NA_character_),
    year     = suppressWarnings(as.integer(col(exp_raw, "year") %||% NA)),
    abstract = NA_character_
  )
  nrow(exp_tidy)
}
```

## Combine and deduplicate

```{r}
#| label: combine

tidy_objs <- c("oa_tidy", "epmc_tidy", "pm_tidy", "s2_tidy", "exp_tidy")
records <- bind_rows(mget(tidy_objs[vapply(tidy_objs, exists, logical(1))]))

records <- records |>
  mutate(
    doi_key   = na_if(doi, ""),
    title_key = na_if(str_squish(str_replace_all(str_to_lower(title), "[^\\p{L}\\p{N}]+", " ")), ""),
    match_key = coalesce(doi_key, title_key)
  )

identified_by_source <- count(records, source, name = "n_identified")

# Prefer the copy that carries an abstract, so the methods flag can see it.
deduped <- records |>
  mutate(has_abs = !is.na(abstract) & abstract != "") |>
  arrange(match_key, desc(has_abs), source) |>
  distinct(match_key, .keep_all = TRUE) |>
  select(-has_abs)

n_identified  <- nrow(records)
n_after_dedup <- nrow(deduped)
n_duplicates  <- n_identified - n_after_dedup
```

## Initial findings

```{r}
#| label: counts

identified_by_source

tibble::tibble(
  metric = c("Records identified (all sources)", "Duplicates removed",
             "Records after dedup"),
  n      = c(n_identified, n_duplicates, n_after_dedup)
)
```

1126 dedup

How many of the deduplicated records mention a psychometric method in their title
or abstract, which approximates the reach of the methods-restricted Search B.
Abstracts are available from OpenAlex and Europe PMC. PubMed and Semantic Scholar
contribute titles only, so this remains a mild lower bound.

```{r}
#| label: methods-share

deduped |>
  mutate(mentions_method = str_detect(paste(title %||% "", abstract %||% ""),
                                      methods_pattern)) |>
  count(mentions_method)
```

187 mentions method.

Publication-year distribution of the deduplicated set.

```{r}
#| label: year-plot

deduped |>
  filter(!is.na(year)) |>
  count(year) |>
  ggplot(aes(year, n)) +
  geom_col() +
  labs(x = "Publication year", y = "Records (deduplicated)") +
  theme_minimal()
```

```{r}
#| label: year-plot-psym

deduped |>
  filter(!is.na(year)) |>
  mutate(mentions_method = str_detect(paste(title %||% "", abstract %||% ""),
                                      methods_pattern)) |>
  filter(mentions_method == TRUE) |> 
  count(year,mentions_method) |> 

  ggplot(aes(year, n)) +
  geom_col(fill = "sienna2") +
  labs(x = "Publication year", y = "Records (deduplicated)",
       subtitle = "Psychometric in title") +
  theme_minimal()
```

## Source contribution and overlap

How much each source overlaps the others, and how many records it uniquely
contributes. This is the direct evidence for the open-only decision and for the
Information Sources section of the protocol: a source that adds few unique
records is dispensable, while a large unique contribution justifies keeping it.

```{r}
#| label: source-contribution

# One row per (record, source) pair.
key_source <- distinct(records, match_key, source)

# Records found in 1, 2, 3, ... sources.
overlap <- key_source |>
  count(match_key, name = "n_src") |>
  count(n_src, name = "n_records")

# Records unique to a single source, tallied by source.
unique_by_source <- key_source |>
  add_count(match_key, name = "n_src") |>
  filter(n_src == 1) |>
  count(source, name = "n_unique")

# Distinct records each source holds, and how many are unique to it.
source_contribution <- key_source |>
  count(source, name = "n_records") |>
  left_join(unique_by_source, by = "source") |>
  mutate(n_unique   = coalesce(n_unique, 0L),
         pct_unique = round(100 * n_unique / n_records, 1)) |>
  arrange(desc(n_records))

overlap
source_contribution
```

  source           n_records n_unique pct_unique
  <chr>                <int>    <int>      <dbl>
1 OpenAlex              1015      264       26  
2 Semantic Scholar       811       88       10.9
3 Europe PMC             425       10        2.4
4 PubMed                 403        7        1.7

```{r}
# The records PubMed contributed that no other source held.
pm_only <- key_source |>
  add_count(match_key, name = "n_src") |>
  filter(n_src == 1, source == "PubMed") |>
  left_join(distinct(records, match_key, title, year, doi), by = "match_key")

pm_only |> select(year, doi, title) |> print(n = Inf) |> 
  knitr::kable()
```

PubMed contributed nothing real. Three of 7 had missing DOI, two had missing title, and the remaining two had no relevance.

## Known-item validation

You need five to ten PSWQ psychometric papers you already know, then check both whether Search A returned them and whether the methods pattern flags them. A known paper that Search A misses means a string revision and a bump to v2 in the search log. A known paper that Search A returns but the methods pattern misses means the methods block has a hole, which matters less since Search B does not decide inclusion.

```{r}
# Known-item check against the pilot pulls. Paste DOIs of 5 to 10 PSWQ
# psychometric papers you already know should be found.
known <- c(
  "10.1037/a0037251",
  "10.1016/j.ijchp.2025.100579",
  "10.1080/02646838.2023.2209101",
  "10.1007/s10862-025-10200-4",
  "10.1007/s10578-024-01774-3",
  "10.4103/2542-3932.304944",
  "10.1007/s10862-023-10029-9",
  "10.1177/1073191121993223",
  "10.1177/10731911211013911",
  "10.1080/03069885.2021.1897969"
)

library(dplyr); library(stringr)
norm <- function(x) str_remove(str_to_lower(x), "^https?://(dx\\.)?doi\\.org/")

# Search A reach
tibble(doi = known) |>
  mutate(found_in_A = norm(doi) %in% norm(deduped$doi)) |> 
  arrange(doi)

# Search B reach: a known paper found by A but not flagged by the methods
# pattern means the methods block would miss it. Needs the abstract column.
deduped |>
  filter(norm(doi) %in% norm(known)) |>
  mutate(mentions_method = str_detect(paste(title, abstract), methods_pattern)) |>
  select(doi, title, mentions_method) |> 
  arrange(doi)
```

All 10 found using A and B. Check was run on the searches made September 8th 2026.

```{r}
# Do any records lack both a DOI and a usable title?
records |>
  filter(is.na(match_key) | match_key == "") |>
  count(source)

sum(is.na(records$match_key)  | records$match_key  == "")   # expect 0
sum(is.na(deduped$match_key)  | deduped$match_key  == "")   # expect 0

records |>
  filter(is.na(match_key) | match_key == "") |>
  mutate(has_abstract = !is.na(abstract) & abstract != "") |>
  select(source, year, has_abstract, doi, title) |>
  print(n = 40)
```


## Hand off to screening

Write the deduplicated set to CSV for import into Rayyan, and keep the full
record table as an `.rds` for later steps.

```{r}
#| label: export

readr::write_csv(
  deduped |> select(source, year, title, doi),
  file.path(cache_dir, "records_deduped.csv")
)
saveRDS(deduped, file.path(cache_dir, "records_deduped.rds"))
```

## Notes

- Each source is cached under `data/`. Delete a file or pass `refresh = TRUE` to
  re-fetch.
- OpenAlex and Europe PMC (via `resultType=core`) return abstracts, so the
  methods-mention flag covers them. PubMed and Semantic Scholar contribute titles
  only here, so the flag stays a mild lower bound.
- The programmatic pulls use Search A (instrument only). Search B, the
  methods-restricted query, is documented per database in Appendix A of
  `index.qmd`, and its OpenAlex reach is approximated above by the
  methods-mention flag.
